Resolvable Ambiguity

### 2019-06-29 paper/paper.tex
EOD

When constructing domain-specific programming languages there is often a desire to reuse work, to compose smaller language fragments. However, how to make such compositions well-behaved is not always obvious. In particular, composing multiple syntaxes is likely to produce an ambiguous grammar, even if all of the original grammars are unambiguous. Approaches for solving this problem can be broadly split into two categories: those that handle ambiguities on the grammar-level (detection, prevention, etc.), and those that work on particular examples of ambiguous words. i.e., ambiguous programs.

>>>>

When constructing domain-specific programming languages there is often a desire to reuse work, to compose smaller language fragments. However, how to make such compositions well-behaved is not always obvious. In particular, composing multiple syntaxes is likely to produce an ambiguous grammar, even if all of the original grammars are unambiguous. Approaches for solving this problem can be broadly split into two categories: those that handle ambiguities on the grammar-level (detection, prevention, etc.), and those that work on particular examples of ambiguous words. i.e., ambiguous programs. The former is well explored in the form of heuristics for ambiguity detection (e.g. \cite{bastenAmbiguityDetectionProgramming2011,axelssonAnalyzingContextFreeGrammars2008,brabrandAnalyzingAmbiguityContextFree2007}) or restrictions on the grammars to be composed (e.g. \cite{kaminskiModularWellDefinednessAnalysis2013}), while the latter has received relatively little scrutiny. Here, previous work is limited to merely presenting each possible interpretation as an abstract syntax tree of some form \cite{palmkvistCreatingDomainSpecificLanguages2019,danielssonParsingMixfixOperators2011}. However, an abstract syntax tree is essentially an implementation detail, and may not provide any useful clues for an end-user to resolve the ambiguity. Another possibility, mentioned by \citet{palmkvistCreatingDomainSpecificLanguages2019}, is to automatically generate several unambiguous programs, each of which produces the same abstract syntax tree as one of the posssible interpretations, thereby directly showing a user how to resolve the ambiguity, regardless of their desired interpretation. For example, in a language with out precedence, ''$1 + 2 * 3$'' is ambiguous, but the two possible interpretations can be written unambiguously as ''$(1 + 2) * 3$'' and ''$1 + (2 * 3)$'' respectively.

In exploring this approach, we find that this is not always possible; some ambiguities have one or more interpretations for which no unambiguous program is exists. This presents a problem, since such an ambiguity is an error an end-user might encounter, yet it can only be solved by the language designer, since it requires a change to the language grammar. We thus have a concept of ambiguities that can be resolved by an end-user, and those that cannot. We call the former \emph{resolvable ambiguities}, and the latter \emph{unresolvable ambiguities}.
